---
title: "Mass Spectrometry Lemierre - Supplementary plots"
author: "Gustav Torisson"
format: pdf
editor: visual
---

```{r, echo = F, message = F, warning=F}
library(here)
library(ggpubr)
library(kableExtra)

source(here("code", "main_script_revision.R"))
source(here("code", "plots.R"))

```

### Descripton of raw data

-   Raw data consists of a wide dataframe with 654 rows with protein names and `r ncol(raw_data)` columns (1 with protein names and 23 samples).
-   These were grouped into 8 with Lemierre syndrome ("LS"), 15 other sepsis ("Sepsis").
-   There was one measurement per sample.

### Initial data management

-   Proteins with several names, separated with semicolon(;) were renamed to only the first name (left of semicolon)
-   Datapoints labelled as "Filtered" were re-labelled as NA.
-   Data was converted to "numeric" format, as it was "character" from the Excel import
-   Datapoints labelled as Nan (Not a Number) were also labelled as NA

\newpage

### Log2 transformation

-   all measurements were log2 transformed

```{r, echo = F, message = F, fig.width= 10, fig.height=5}
#| fig-cap: "Density plots of raw and log2 transformed values"
#| fig-cap-location: bottom

print(ggarrange(raw_plot, log_plot, nrow = 1))

```

\newpage

### Filtering

-   Of 654 proteins, 341 (52.1%) had complete data, in all 23 samples
-   205 proteins (31.3%) were missing in \>= 7 (30% of all) samples.
-   These were filtered, leaving 449 proteins

```{r, echo = F, message = F, fig.width = 10, fig.height=5}
#| fig-cap: "Missing structure before and after filterina"
#| fig-cap-location: bottom

print(ggarrange(plot_before_filt, plot_after_filt, nrow = 1))

```

\newpage

### Normalisation

-   all values were normalised by sample by subtracting the sample median from the Log2 intensity values

```{r, echo = F, message = F, fig.width = 10, fig.height=5, fig.width= 8  }
#| fig-cap: "Before and after by-sample normalisation."

print(ggarrange(before_norm_plot, after_norm_plot, nrow = 1))

```

\newpage

### Missing per sample and group

-   All NAs were considered to represent low intensities and to represent MNAR (missing not at random)

```{r, echo = F, message = F, fig.width = 5}
#| fig-cap: "Valid measurements after filtering per sample and group"
print(valid_plot)

```

\newpage

### Imputation

-   NAs were imputed using single imputation, assuming MNAR
-   For each sample, the sample mean and sample sd were determined
-   Then imputations were performed, using a random draw from a Gaussian distribution
-   The mean for imputations was donwshifted with -1.8 sample sd and the width 0.3 \* sample sd

```{r, echo =FALSE, message = FALSE, fig.height=4}
#| fig-cap: "Distribution of imputed values."

print(impute_plot)

```

\newpage

### Differential expression

-   449 t-tests (students t-test) were performed, one-at-a-time for each protein, between LS and Sepsis groups
-   Results are presented as:
    -   Log2FC = mean(log(LS)) - mean(log(Sepsis))
    -   p values from t-test
    -   q values using Benjamini-Hochberg corrections
    -   FC(Fold change) = 2\^Log2FC
    -   Values with Log2FC ± 1.0 and q value \< 0.05 were considered significant

```{r, echo = F, message = F}
#| tbl-cap: "Differential expression between LS and Sepsis."
kable(table1, format = "markdown")

```

\newpage

### Volcano plot

```{r, echo = F, message = F, fig.width=4}
#| fig-cap: "Volcano plot. Proteins that are differentially expressed between LS and Sepsis = red"

print(volcano_plot)
```

\newpage

### Heatmap

```{r, echo = F, message = F, fig.width=6.5}
#| fig-cap: heatmap of differentially expressed proteins between LS and Other severe infections
print(heat_plot)


```

\newpage
```{r, echo = F, message = F}
#| tbl-cap: Table underlying the heatmap above. LS and Sepsis are mean normalized log2-transformed intensities
kable(heat_plot_table, format = "markdown")

```
