---
title: "Mass Spectrometry Lemierre - Working document"
author: "Gustav Torisson"
format: pdf
editor: visual
---

```{r, echo = F, message = F, warning=F}
library(here)
library(ggpubr)
library(kableExtra)

source(here("code", "new_script.R"))

```

### Descripton of raw data

-   Raw data consists of a wide dataframe with 654 rows with protein names and 30 columns (1 with protein names and 29 samples.
-   These were grouped into 8 with Lemierre syndrome ("LS"), 15 other sepsis ("Sepsis"), 3 Deep vein thrombosis ("DVT") and 3 other septic thromboses ("Other_thrombosis").
-   There was one measurement per sample.
-   Each data point represents an intensity measure from Mass pectrometry.

### Initial data management

-   Proteins with several names, separated with semicolon(;) were renamed to only the first name (left of semicolon)
-   Datapoints labelled as "Filtered" were re-labelled as NA.
-   Data was converted to "numeric" format, as it was "character" in Excel
-   Datapoints labeeled as Nan (Not a Number) were also labelled as NA

\newpage

### Log2 transformation

-   all measurements were log2 transformed

```{r, echo = F, message = F, fig.width= 10, fig.height=5}
#| fig-cap: "Density plots of raw and log2 transformed values"
#| fig-cap-location: bottom

print(ggarrange(raw_plot, log_plot, nrow = 1))

```

\newpage

### Filtering

-   Of 654 proteins, 329 (50.3%) had complete data, in all 29 samples
-   200 proteins (30.6%) were missing in \> 9 (30% of all) samples.
-   These were filtered, leaving 454 proteins

```{r, echo = F, message = F, fig.width = 10, fig.height=5}
#| fig-cap: "Missing structure before and after filterina"
#| fig-cap-location: bottom

print(ggarrange(plot_before_filt, plot_after_filt, nrow = 1))

```

\newpage

### Normalisation

-   all values were normalised by sample by subtracting the sample median from the Log2 intensity values

```{r, echo = F, message = F, fig.width = 10, fig.height=5, fig.width= 8  }
#| fig-cap: "Before and after by-sample normalisation."

print(ggarrange(before_norm_plot, after_norm_plot, nrow = 1))

```

\newpage

### Missing per sample and group

-   there were most missing values in the DVT group, but differences were small
-   All NAs (both "Filtered" and NaN) were considered to represent low intensities and to represent MNAR (Correct???)

```{r, echo = F, message = F, fig.width = 5}
#| fig-cap: "Valid measurements after filtering per sample and group"
print(valid_plot)

```

\newpage

### Imputation

-   NAs were imputed using single imputation, assuming MNAR
-   For each sample, the sample mean and sample sd were determined
-   Then imputations were performed, using a random draw from a Gaussian distribution
-   The mean for imputations was donwshifted with -1.8 sample sd and the width 0.3 \* sample sd

```{r, echo =FALSE, message = FALSE, fig.height=4}
#| fig-cap: "Distribution of imputed values."

print(impute_plot)

```

\newpage

### Differential expression

-   454 t-tests (students t-test) were performed, one for each protein, between LS and Sepsis groups
-   Results are presented as:
    -   Log2FC = mean(log(LS)) - mean(log(Sepsis))
    -   p values from t-test
    -   q values using Benjamini-Hochberg corrections
    -   FC(Fold change) = 2\^Log2FC
    -   Values with Log2FC ± 1.0 and q value \< 0.05 were considered significant
    -   number of imputed NAs (in LS and Sepsis groups)

```{r, echo = F, message = F}
#| tbl-cap: "Differential expression between LS and Sepsis."
kable(table1, format = "markdown")

```

\newpage

### Including LS vs VTE

-   the proteins that were differentially expressed for LS vs Sepsis were also evaluated for differences between LS and VTE.

```{r, echo = F}
#| tbl-cap: "LS vs Sepsis (left, FC1, q1 and sign1) and LS vs VTE (right, FC2, q2, sign2)"
kable(table3, format = "markdown")

```

\newpage

### Volcano plot

```{r, echo = F, message = F, fig.width=4}
#| fig-cap: "Volcano plot. Proteins that are differentially expressed between LS and Sepsis = red"

print(volcano_plot)
```

### Heatmap

```{r, echo = F, message = F}
print(heat_plot)

```

\newpage

### KOMMENTARER TILL DAVID

-   Jag fick 23 protein jmfrt med 25 i manuset. det är 3 av dina som inte är med (mer om det nedan) och sen har det tillkommit ett (A0A0C4DH33), detta har qval precis \< 0.05 och 5 imputationer så rätt känsligt för variationer i imputationen
-   Jag gjorde volcano plotten med -log10 q värde istället för pvärde (känns mer rätt eftersom man kan sätta strecket på 1.3 för -log10), därför de skiljer sig på y-axeln.
-   Nedan finns en tabell med de 25 proteiner som var signifikanta i din körning, de är sorterade i bokstavsordning
-   3 st är inte med här men var med i manuset, det är
    -   P05556 - som verkar imputationskänslig, varierar väldigt mkt
    -   P07998 och Q08380 - som ligger väldigt nära 0.05 och därför blir känsliga för variation

```{r, echo = F}
kable(david_tbl, format = "markdown")
```
